---
title: "Donsker's Thoerem Ideas"
pkgdown:
  as_is: true
output:
  bookdown::html_document2:
  number_sections: true
toc: true
toc_depth: 3
toc_float:
  collapsed: false
header-includes:
  - \usepackage{breqn}
  - \usepackage{palatino}
  - \usepackage{xcolor}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)

quick_eval <- FALSE
```

```{r libraries, include = FALSE, warning = FALSE, message = FALSE}
library(MASS)
library(mvtnorm)
library(latex2exp)
```


# Donsker's Theorem

Let $F_n$ be the empirical distribution function of i.i.d. random variables $X_1, X_2, X_3,...$ with distribution function $F$. Define the centered and scaled version of $F_n$ by

\[
G_n(x) = \sqrt{n}(F_n(x) - F(x))
(\#eq:donskerseq)
\]

indexed by $x \in \mathbb{R}$. From the classical central limit theorem for fixed $x$, we know the random variable $G_n(x)$ converges in distribution to a Gaussian random variable with mean zero and variance $F(x)(1 - F(x))$ as the sample size $n$ grows. Donsker's theorem adds to this, and shows that $G_n(x)$ converges in distribution to a Gaussian Process with mean zero and covariance $K(G(s), G(t))$ given by 

\[
K(G(s), G(t)) = E[G(s)G(t)] = min\{F(s), F(t)\} - F(s)F(t)
(\#eq:donskerkernel)
\]

## Continuous Example

Let $x \sim N(0, 1)$. 

First, lets look at some potential CDFs from sampling $X_n$ with $n = 1000$.

```{r normexample, echo = FALSE, out.width = "100%", fig.cap = "Empirical sample CDFs for $X$"}
N <- 50
n <- 1000
storage <- matrix(rnorm(N * n), nrow = N, ncol = n)

colors <- sample(colors()[-1], N)
plot(NA, NA, xlim = c(-3, 3), ylim = c(-0.05, 1.05), xlab = "x", ylab = "F(x)")
for(i in 1:N) {
  lines(sort(storage[i, ]), (1:n) / n, type = "s", col = colors[i])
}
```

Next, lets transform these $F_n(x)$, using Donsker's theorem.

```{r donskerize}
storage2 <- matrix(NA, nrow = N, ncol = n)

for(i in 1:N) {
  storage2[i, ] <- sqrt(n) * (sapply(storage[i, ], function(x) sum(storage[i, ] <= x)) / n - 
                                pnorm(storage[i, ]))
}
```

And now we can visualize them.

```{r donskervisual, echo = FALSE, out.width = "100%", fig.cap = "Transformed CDF values"}
plot(NA, NA, xlim = c(-3, 3), ylim = c(-1.5, 1.5), xlab = "x", ylab = TeX("$\\sqrt{n}(F_n(x) - F(x))$"))
for(i in 42:42) {
  ord <- order(storage[i, ])
  lines(storage[i, ord], storage2[i, ord], type = "l", col = colors[i])
}
```

Extracting densities.

```{r donskerdensities}
covMat <- function(p){
  outer(p, p, FUN = "pmin") - outer(p, p)
}

denStore <- rep(NA, N)

for(i in 1:N) {
  ord <- order(storage[i, ])
  K <- covMat(pnorm(storage[i, ord]))[-n, -n]
  
  test <- rmvnorm(1, sigma = K)
  dmvnorm(test, sigma = K, log = TRUE)
  
  denStore[i] <- dmvnorm(sqrt(n) * (sapply(storage[i, ord], function(x) sum(storage[i, ] <= x)) / n - pnorm(storage[i, ord]))[-n], sigma = K, log = TRUE)
}

denStore
```

## Discrete Example

Let $x \sim Unif(\{1, 2, 3, \dots, 100\})$. 

First, lets look at some potential CDFs from sampling $X_n$ with $n = 1000$.

```{r discexample, echo = FALSE, out.width = "100%", fig.cap = "Empirical sample CDFs for $X$"}
N <- 10
n <- 1000
storage <- matrix(sample(1:100, N * n, replace = TRUE), nrow = N, ncol = n)

plot(NA, NA, xlim = c(0, 101), ylim = c(-0.05, 1.05), xlab = "x", ylab = "F(x)")
for(i in 1:N) {
  FX <- sapply(1:100, function(x) sum(storage[i, ] <= x)) / n
  lines(1:100, FX, type = "s", col = colors[i])
}
```

Next, lets transform these $F_n(x)$, using Donsker's theorem.

```{r donskerize2}
storage2 <- matrix(NA, nrow = N, ncol = 100)

for(i in 1:N) {
  storage2[i, ] <- sqrt(n) * (sapply(1:100, function(x) sum(storage[i, ] <= x) / n - x / 100))
}
```

And now we can visualize them.

```{r donskervisual2, echo = FALSE, out.width = "100%", fig.cap = "Transformed CDF values"}
plot(NA, NA, xlim = c(0, 101), ylim = c(-1.5, 1.5), xlab = "x", ylab = TeX("$\\sqrt{n}(F_n(x) - F(x))$"))
for(i in 1:N) {
  lines(1:100, storage2[i, ], type = "l", col = colors[i])
}
```

Extracting densities.

```{r donskerdensities2}
covMat <- function(p){
  outer(p, p, FUN = "pmin") - outer(p, p)
}

denStore <- rep(NA, N)

for(i in 1:N) {
  K <- covMat(1:100 / 100)[-100, -100]
  
  # test <- rmvnorm(1, mean = rep(0, 99), sigma = K[-100, -100])
  # dmvnorm(test, sigma = K[-100, -100])
  
  denStore[i] <- dmvnorm(storage2[i, -100], sigma = K)
}

denStore
```
