---
title: "Ideas"
pkgdown:
  as_is: true
output:
  bookdown::html_document2:
    number_sections: true
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
header-includes:
  - \usepackage{breqn}
  - \usepackage{palatino}
  - \usepackage{xcolor}
---

```{r setup, include = FALSE}
library(dplyr)
library(ggplot2)
library(gridExtra)

knitr::opts_chunk$set(
  cache = TRUE,
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)

quick_eval <- FALSE
```

# Assumptions

## Membership in windows is uniform

If $X_n$ is an ordered sample from a population $Y_N$ with $n < N$, then the *average* behavior of $Y_N \setminus X_n$ can be described by $Pr[y_{k} \in (z_i, z_{i+1})]\ = \frac{1}{n+1} \forall\ k \in n+1, \dots, N\ \wedge\ i \in 1,\dots,n+1$ where $Z = \{-\infty, x_1, \dots, x_n, \infty\}$.

### Simulation

```{r}
iters <- 50000
N <- 500
n <- 20

pop <- runif(N)

storage <- matrix(NA, nrow = iters, ncol = n + 1)

for(i in 1:iters) {
  ind <- sample(1:N, n)
  samp <- sort(pop[ind])
  others <- pop[-ind]
  
  tabs <- table(findInterval(pop, c(-Inf, samp, Inf), left.open = TRUE))
  add1 <- as.numeric(tabs)
  names(add1) <- names(tabs)
  add2 <- rep(0, n + 1)
  names(add2) <- 1:(n + 1)
  add <- list(add1, add2)

  tab <- tapply(unlist(add), names(unlist(add)), sum)
  storage[i, ] <- tab[order(factor(names(tab), levels = 1:(length(samp) + 1)))]
}
```

```{r, echo = FALSE, out.width = "100%", fig.height = 3, fig.cap = "Membership of $Y_N \\setminus X_n$ for a single sample (left), and average behavior over many samples (right), with error bars defined using 10th and 90th percentiles"}
g1 <- data.frame(
  "Window" = 1:(n + 1),
  "Membership" = storage[1, ]
) %>%
  ggplot(mapping = aes(y = as.factor(Window), x = Membership)) + 
  geom_bar(stat = "identity") + 
  labs(x = "Membership", y = "Window")

g2 <- data.frame(
  "Window" = 1:(n + 1),
  "Mean" = colMeans(storage),
  "Q10" = apply(storage, 2, quantile, probs = 0.1),
  "Q90" = apply(storage, 2, quantile, probs = 0.9)
) %>%
  ggplot(mapping = aes(y = as.factor(Window), x = Mean)) + 
  geom_bar(stat = "identity") + 
  geom_errorbar(mapping = aes(xmin = Q10, xmax = Q90), color = "red") + 
  labs(x = "Mean Membership", y = "Window")

grid.arrange(g1, g2, nrow = 1)
```

### Reversibility

If we instead fix our sample, and create synthetic populations using the same distribution that generated our original population, does this property apply to the synthetic populations?

```{r}
iters <- 50000
N <- 500
n <- 20

orig.pop <- runif(N)
samp <- sort(sample(orig.pop, n))

storage2 <- matrix(NA, nrow = iters, ncol = n + 1)

for(i in 1:iters) {
  pop <- runif(N - n)
  
  tabs <- table(findInterval(pop, c(-Inf, samp, Inf), left.open = TRUE))
  add1 <- as.numeric(tabs)
  names(add1) <- names(tabs)
  add2 <- rep(0, n + 1)
  names(add2) <- 1:(n + 1)
  add <- list(add1, add2)

  tab <- tapply(unlist(add), names(unlist(add)), sum)
  storage2[i, ] <- tab[order(factor(names(tab), levels = 1:(length(samp) + 1)))]
}
```

```{r}
g1 <- data.frame(
  "Window" = 1:(n + 1),
  "Mean" = colMeans(storage2),
  "Q10" = apply(storage2, 2, quantile, probs = 0.1),
  "Q90" = apply(storage2, 2, quantile, probs = 0.9)
) %>%
  ggplot(mapping = aes(y = as.factor(Window), x = Mean)) + 
  geom_bar(stat = "identity") + 
  geom_errorbar(mapping = aes(xmin = Q10, xmax = Q90), color = "red") + 
  labs(x = "Mean Membership", y = "Window")

grid.arrange(g1, nrow = 1)
```

