---
title: "Population Synthesis"
pkgdown:
  as_is: true
output:
  bookdown::html_document2:
    keep_tex: true
    number_sections: true
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  eval = TRUE,
  cache = TRUE
)

quick_eval <- FALSE
options(scipen = 6)
set.seed(318937291)
```

```{r libraries, include = FALSE, warning = FALSE, message = FALSE}
library(extraDistr)
library(lemur.pack)
library(MASS)
library(mvtnorm)
library(latex2exp)
library(microbenchmark)
library(tidyverse)
library(gridExtra)
library(cowplot)
```

# Populations

We are working with various finite populations which take on different values, ranging from simple discrete-valued categorical variables, to continuous-valued variables. We focus on three scenarios:

(1) Categorical -- $Y_i \in \{a, b, c, \dots\}$
(2) Discrete Numeric -- $Y_i \in \{1, 2, 3, \dots\}$
(3) Continuous Numeric -- $Y_i \in \mathbb{R}$

# Synthesis

We employ *Markov chain Monte Carlo* (MCMC) methods to sample from the posterior of our populations, and focus largely on jittering populations via symmetric proposals. Our posterior is constructed the canonical way, via

\[
\pi(y|x) \propto \ell(x|y) p(y)
\]

The two major components, our prior distribution $p(y)$, and likelihood function $\ell(x|y)$.

## Likelihood

Because we are working with finite populations, $\ell(x|y)$ often cannot take on the form of sampling distribution. To illustrate, imagine a population...

[insert example here]

## Prior

[still need to talk about priors... but we have the discrete case]




# Comparisons

## Categorical Populations

### Binary

#### Binomial vs Hypergeometric

#### Brownian Bridge using Integers

### Many Categories

#### Multinomial vs Multivariate Hypergeometric

##### Problems as number of categories increases

#### Brownian Bridge using Integers (cont'd)

#### Dirichlet Approx using Integers

### Discrete Numeric Populations

#### Problems with Multinomial & Multivariate Hypergeometric

#### Brownian Bridge

#### Dirichlet Approx

#### Problems as number of unique values increases? -> Motivation for fixed interpolation points

### Continuous Populations (bounded)

#### Brownian Bridge

#### Interpolated

#### Dirichlet Approx

##### Equal probability on endpoints vs 1/2 probability on 

### Continuous Populations (unbounded)

???

### Parametric Populations

#### Normal "lock-step" fixed intervals in inverse-cdf space

#### Changing from "lock-step" -> i.i.d.

### Binned data informing likelihood




