---
title: "Population Representativeness"
pkgdown:
  as_is: true
output:
  bookdown::html_document2:
  keep_tex: true
number_sections: true
toc: true
toc_depth: 3
toc_float:
  collapsed: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  eval = TRUE,
  cache = TRUE
)

quick_eval <- FALSE
options(digits = 2)
options(scipen = 6)
set.seed(318937291)
```

```{r libraries, include = FALSE, warning = FALSE, message = FALSE}
library(extraDistr)
library(lemur.pack)
library(MASS)
library(mvtnorm)
library(latex2exp)
library(microbenchmark)
library(tidyverse)
library(gridExtra)
library(cowplot)
```

# Introduction

::: {.definition #representative}
In statistics, the term *representative* is commonly applied to a sample that meets either of two conditions conditions, 1) sampling was conducted in a way that leads to each possible sample being equally likely, and 2) it is possible to draw accurate conclusions about a population from the sample.
:::

While the second condition may imply that the first condition was met (it is certainly *not* guaranteed), it would be foolish to imply that the first condition would imply the second. For example, consider a population $Y$ that is the first $N=99$ natural numbers: $1, 2, \dots, 99$. Suppose two i.i.d. samples of size $n = 9$, $x_1$ and $x_2$ are both taken from this population.

```{r, echo = FALSE}
Y <- 1:99
x_1 <- c(1, 3, 7, 34, 37, 71, 83, 90, 91)
x_2 <- c(10, 20, 30, 40, 50, 60, 70, 80, 90)
```


\begin{align*}
  x_1 &= \{1, 3, 7, 34, 37, 71, 83, 90, 91\} \\
  x_2 &= \{10, 20, 30, 40, 50, 60, 70, 80, 90\}
\end{align*}

In what way is $x_2$ more *representative* of the population $Y$? For starters, the mean of $Y$ is `r mean(Y)`, while the mean of $x_1$ and $x_2$ are `r mean(x_1)` and `r mean(x_2)` respectively. Likewise, the standard deviations are `r sd(Y)` for the population, `r sd(x_1)`, and `r sd(x_2)` for the samples. There is another way that $x_2$ is a more desirable sample. Consider the number of population members in each of the $k = n + 1$ spaces defined by $(-\infty, x_{(1)}), (x_{(2)}, x_{(1)}), \dots, (x_{(n)}, \infty)$. The count for $x_2$ is constant; there are 9 population members in each of the 10 spacings. Thus, using this way of comparing samples, $x_2$ would be the ideal sample of $Y$, at least when restricting to samples of size $n = 9$.

Imagine we repeat this sampling process (with replacement) repeatedly, and calculate the number of population members within each space each time.

```{r}
n_samples <- 10000
n <- 9
storage <- matrix(NA, nrow = n_samples, ncol = n + 1)

for(i in 1:n_samples) {
  samp <- sort(sample(Y, size = n))
  spacings <- tabulate(findInterval(Y, samp, left.open = TRUE) + 1, nbins = n + 1) - c(rep(1, n), 0)
  storage[i, ] <- spacings
}

apply(storage, 2, mean)
apply(storage, 2, sd)
```

