---
title: "Sampler Comparisons"
pkgdown:
  as_is: true
output:
  bookdown::html_document2:
    number_sections: true
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
header-includes:
  - \usepackage{breqn}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)

quick_eval <- FALSE
```

```{r paramset, echo = FALSE, eval = TRUE}
if(quick_eval) {
  samples <- 500
} else {
  samples <- 20000
}

set.seed(1234)
```

```{r packageload, message = FALSE, warning = FALSE, eval = TRUE}
library(extraDistr)
library(knitr)
library(kableExtra)
library(proftools)
```

# Intro

Text here.

# Samplers

Here we will compare three samplers:

## Importance Sampling

```{r importanceSampler}
importanceSampler <- function(x, N, samples = 1000, saturation = 10, nEff = length(sample), 
                              alpha = 0.5, cutoff = 0.5) {
  n <- length(x)
  nLow <- sum(x < cutoff)

  priorDrawsRaw <- rdirichlet(samples * saturation, (nEff + alpha) / (N + 1) * rep(1, N + 1))
  priorDraws <- t(apply(matrix(priorDrawsRaw, nrow = samples * saturation)
                        [, -(N + 1), drop = FALSE], 1, cumsum))
  NLow <- apply(priorDraws, 1, function(x) sum(x < cutoff))
  weights <- dhyper(nLow, NLow, N - NLow, n)

  draws <- sample(1:(samples * saturation), size = samples, replace = TRUE, prob = weights)

  return(priorDraws[draws, ])
}
```

## MCMC with Uniform Proposal

```{r mcmcUniform}
mcmcUniform <- function(x, N, samples = 1000, nEff = length(sample), alpha = 0.5, cutoff = 0.5) {

  lpost <- function(nLow, pop, N, nEff, alpha) {
    NLow <- sum(pop < cutoff)
    diffs <- diff(c(0, sort(pop), 1))
    lprior <- ddirichlet(diffs, (nEff + alpha) / (N + 1) * rep(1, N + 1), log = TRUE)
    llik <- dhyper(nLow, NLow, N - NLow, n, log = TRUE)
    return(llik + lprior)
  }

  n <- length(x)
  nLow <- sum(x < cutoff)

  out <- matrix(NA, nrow = samples, ncol = N)
  # out[1, ] <- rep(x, length.out = N) + runif(N, 0, 0.01)
  out[1, ] <- runif(N)
  lp0 <- lpost(nLow, out[1, ], N, nEff, alpha)

  for(i in 2:samples) {
    current <- out[i - 1, ]
    for(k in 1:N) {
      proposal <- current
      proposal[k] <- runif(1)
      lp1 <- lpost(nLow, proposal, N, nEff, alpha)

      u <- runif(1)
      if(is.finite(lp1) && log(u) < (lp1 - lp0)) {
        current <- proposal
        lp0 <- lp1
      }
    }
    out[i, ] <- current
  }
  return(out)
}
```


## MCMC with Conditional Prior Proposal

Let $x_1, x_2, \dots, x_N$ represent the *ordered* population of size *N*. We denote $s_1 = x_1$, $s_2 = x_2 - x_1, \dots, s_{N-1} = x_N - x_{N-1}$, $s_N = 1 - x_N$ as the spacing of the ordered population.

From our formulation of the posterior, $\sum s_i = 1$ and $(s_1, \dots, s_{N+1}) \sim Dir(\frac{N_E + \alpha}{N+1}, \dots, \frac{N_E + \alpha}{N+1})$. We can use this to figure out the conditional prior for a set of spacings $s_{(m)}$.

For example, suppose we want the distribution of $x_{(1)} | x_{(2)}$, where $x_{(1)}, x_{(2)}$ partition $x$. Fortunately, this is relatively straightforward, and we end up with

$$ \frac{1}{1 - \mathbf{1}^Tx_{(2)}} X_{(1)} | X_{(2)} = x_{(2)} \sim Dir(\frac{N_E + \alpha}{N+1}, \dots, \frac{N_E + \alpha}{N+1})$$

Essentially, nothing changes except we now know the spacings have to add up to $1 - \mathbf{1}^Tx_{(2)}$. We can use this to our advantage, using this conditional prior as our proposal. For simplicity, we let $N_E\ |\ N$, and consider proposal chunks of size $N / N_E$.

```{r mcmcPrior}
mcmcPrior <- function(x, N, samples = 1000, nEff = length(sample), alpha = 0.5, cutoff = 0.5) {

  lpost <- function(nLow, pop, N, nEff, alpha) {
    NLow <- sum(pop < cutoff)
    diffs <- diff(c(0, sort(pop), 1))
    lprior <- ddirichlet(diffs, (nEff + alpha) / (N + 1) * rep(1, N + 1), log = TRUE)
    llik <- dhyper(nLow, NLow, N - NLow, n, log = TRUE)
    return(llik + lprior)
  }

  n <- length(x)
  nLow <- sum(x < cutoff)

  out <- matrix(NA, nrow = samples, ncol = N)
  # out[1, ] <- rep(x, length.out = N) + runif(N, 0, 0.01)
  out[1, ] <- runif(N)
  lp0 <- lpost(nLow, out[1, ], N, nEff, alpha)

  for(i in 2:samples) {
    current <- out[i - 1, ]
    for(k in 1:N) {
      proposal <- current
      proposal[k] <- runif(1)
      lp1 <- lpost(nLow, proposal, N, nEff, alpha)

      u <- runif(1)
      if(is.finite(lp1) && log(u) < (lp1 - lp0)) {
        current <- proposal
        lp0 <- lp1
      }
    }
    out[i, ] <- current
  }
  return(out)
}
```




# Comparisons

Text here.

```{r samplesSetup}
x1 <- c(0.11, 0.19, 0.27, 0.39, 0.42, 0.48, 0.59, 0.71, 0.84, 0.92)
x2 <- c(0.07, 0.12, 0.19, 0.26, 0.31, 0.36, 0.41, 0.49, 0.67, 0.89)
```

```{r popSetup}
N <- 40
nEff <- 4
alpha <- 0.5
cutoff <- 0.5
```

```{r initRuns, include = FALSE}
initRun1 <- importanceSampler(x = x1, N = N, samples = 100, nEff = 4)
initRun2 <- mcmcUniform(x = x1, N = N, samples = 100, nEff = 4)

initRun3 <- importanceSampler(x = x2, N = N, samples = 100, nEff = 4)
initRun4 <- mcmcUniform(x = x2, N = N, samples = 100, nEff = 4)
```

```{r timings}
pops1_1 <- importanceSampler(x = x1, N = N, samples = samples, nEff = 4, cutoff = 0.6)
pops1_2 <- mcmcUniform(x = x1, N = N, samples = samples, nEff = 4, cutoff = 0.6)

pops2_1 <- importanceSampler(x = x2, N = N, samples = samples, nEff = 4, cutoff = 0.6)
pops2_2 <- mcmcUniform(x = x2, N = N, samples = samples, nEff = 4, cutoff = 0.6)
```

## Location (Mean)

```{r means, echo = FALSE, fig.cap = "Density Comparison of Importance Sampling (red) vs Markov Chain Approach (blue)"}
plot(density(rowMeans(pops1_1)), col = "red",
     main = "", 
     xlab = "Mean of Population")
lines(density(rowMeans(pops1_2)), col = "blue")
```

## Location (CDFs)

```{r cdfs, echo = FALSE, fig.cap = "CDFs of Importance Sampling (red) vs Markov Chain Approach (blue)"}
par(mfrow = c(1, 2))

pops1_1_sort = apply(pops1_1, 2, sort)
plot(c(0, 1), c(0, 1), type = 'l', xlab = 'U', ylab = 'Cumulative Probabilities')
matlines(t(pops1_1_sort), cumsum(rep(1/N, N)), lty = 1, type = 's')
mtext("Importance Sampling")

pops1_2_sort = apply(pops1_2, 2, sort)
plot(c(0, 1), c(0, 1), type = 'l', xlab = 'U', ylab = 'Cumulative Probabilities')
matlines(t(pops1_2_sort), cumsum(rep(1/N, N)), lty = 1, type = 's')
mtext("Markov Chains")
```


