---
title: "Sampler Comparisons"
pkgdown:
  as_is: true
output:
  bookdown::html_document2:
    number_sections: true
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
header-includes:
  - \usepackage{breqn}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)

quick_eval <- FALSE
```

```{r paramset, echo = FALSE, eval = TRUE}
if(quick_eval) {
  samples <- 1000
} else {
  samples <- 20000
}

set.seed(1234)
```

```{r packageload, message = FALSE, warning = FALSE, eval = TRUE}
library(extraDistr)
library(knitr)
library(kableExtra)
library(proftools)
library(dplyr)
```

# Intro

Text here.

# Samplers

Here we will compare three samplers:

## Importance Sampling

```{r importanceSampler}
importanceSampler <- function(x, N, samples = 1000, saturation = 10, nEff = length(sample), 
                              alpha = 0.5, cutoff = 0.5) {
  n <- length(x)
  nLow <- sum(x < cutoff)

  priorDrawsRaw <- rdirichlet(samples * saturation, (nEff + alpha) / (N + 1) * rep(1, N + 1))
  priorDraws <- t(apply(matrix(priorDrawsRaw, nrow = samples * saturation)
                        [, -(N + 1), drop = FALSE], 1, cumsum))
  NLow <- apply(priorDraws, 1, function(x) sum(x < cutoff))
  weights <- dhyper(nLow, NLow, N - NLow, n)

  draws <- sample(1:(samples * saturation), size = samples, replace = TRUE, prob = weights)

  return(priorDraws[draws, ])
}
```

## MCMC with Uniform Proposal

```{r mcmcUniform}
mcmcUniform <- function(x, N, samples = 1000, nEff = length(sample), alpha = 0.5, cutoff = 0.5) {

  lpost <- function(nLow, pop, N, nEff, alpha) {
    NLow <- sum(pop < cutoff)
    diffs <- diff(c(0, sort(pop), 1))
    lprior <- ddirichlet(diffs, (nEff + alpha) / (N + 1) * rep(1, N + 1), log = TRUE)
    llik <- dhyper(nLow, NLow, N - NLow, n, log = TRUE)
    return(llik + lprior)
  }

  n <- length(x)
  nLow <- sum(x < cutoff)

  out <- matrix(NA, nrow = samples, ncol = N)
  # out[1, ] <- rep(x, length.out = N) + runif(N, 0, 0.01)
  out[1, ] <- runif(N)
  lp0 <- lpost(nLow, out[1, ], N, nEff, alpha)

  for(i in 2:samples) {
    current <- out[i - 1, ]
    for(k in 1:N) {
      proposal <- current
      proposal[k] <- runif(1)
      lp1 <- lpost(nLow, proposal, N, nEff, alpha)

      u <- runif(1)
      if(is.finite(lp1) && log(u) < (lp1 - lp0)) {
        current <- proposal
        lp0 <- lp1
      }
    }
    out[i, ] <- current
  }
  return(out)
}
```


## MCMC with Conditional Prior Proposal

Let $x_1, x_2, \dots, x_N$ represent the *ordered* population of size *N*. We denote $s_1 = x_1$, $s_2 = x_2 - x_1, \dots, s_{N-1} = x_N - x_{N-1}$, $s_N = 1 - x_N$ as the spacing of the ordered population.

From our formulation of the posterior, $\sum s_i = 1$ and $(s_1, \dots, s_{N+1}) \sim Dir(\frac{N_E + \alpha}{N+1}, \dots, \frac{N_E + \alpha}{N+1})$. We can use this to figure out the conditional prior for a set of spacings $s_{(m)}$.

For example, suppose we want the distribution of $x_{(1)} | x_{(2)}$, where $x_{(1)}, x_{(2)}$ partition $x$. Fortunately, this is relatively straightforward, and we end up with

$$ \frac{1}{1 - \mathbf{1}^Tx_{(2)}} X_{(1)} | X_{(2)} = x_{(2)} \sim Dir(\frac{N_E + \alpha}{N+1}, \dots, \frac{N_E + \alpha}{N+1})$$

Essentially, nothing changes except we now know the spacings have to add up to $1 - \mathbf{1}^Tx_{(2)}$. We can use this to our advantage, using this conditional prior as our proposal. For simplicity, we let $N_E\ |\ N$, and consider proposal chunks of size $N / N_E$.

```{r mcmcPrior}
mcmcPrior <- function(x, N, samples = 1000, nEff = length(sample), alpha = 0.5, cutoff = 0.5) {

  lpost <- function(nLow, pop, diffs, N, nEff, alpha, cutoff) {
    NLow <- sum(pop < cutoff)
    lprior <- ddirichlet(diffs, (nEff + alpha) / (N + 1) * rep(1, N + 1), log = TRUE)
    llik <- dhyper(nLow, NLow, N - NLow, n, log = TRUE)
    return(llik + lprior)
  }
  
  priorProposal <- function(x) {
    as.numeric(rdirichlet(1, rep((nEff + alpha) / length(x), length(x))) * sum(x))
  }
  
  priorDensity <- function(x) {
    ddirichlet(x / sum(x), rep((nEff + alpha) / length(x), length(x)), log = TRUE)
  }
  
  spacings <- matrix(NA, nrow = samples, ncol = N + 1)
  data <- matrix(NA, nrow = samples, ncol = N)
  
  chunks <- split(1:(N + 1), cut(seq_along(1:(N + 1)), nEff * 2, labels = FALSE))
  
  chunks2 <- list()
  for(i in 1:(length(chunks) - 1)) {
    chunks2[[i]] <- c(chunks[[i]], chunks[[i + 1]])
  }
  
  
  n <- length(x)
  nLow <- sum(x < cutoff)
  
  data[1, ] <- sort(c(runif(ceiling(N * nLow / n), 0, cutoff), 
                      runif(floor(N * (n - nLow) / n), cutoff, 1)))
  spacings[1, ] <- diff(c(0, data[1, ], 1))
  
  
  lp0 <- lpost(nLow, data[1, ], spacings[1, ], N, nEff, alpha, cutoff)
  
  for(i in 2:samples) {
    currentSpacings <- spacings[i - 1,]
    currentData <- data[i - 1, ]
    for(k in 1:length(chunks2)) {
      proposeSpacings <- currentSpacings
      proposeSpacings[chunks2[[k]]] <- priorProposal(proposeSpacings[chunks2[[k]]])
      proposeData <- cumsum(proposeSpacings)[-(N + 1)]
      lp1 <- lpost(nLow, proposeData, proposeSpacings, N, nEff, alpha, cutoff)
      
      u <- runif(1)
      if(is.finite(lp1) && log(u) < (lp1 - priorDensity(proposeSpacings)) - (lp0 - priorDensity(currentSpacings))) {
        currentSpacings <- proposeSpacings
        currentData <- proposeData
        lp0 <- lp1
      }
    }
    spacings[i, ] <- currentSpacings
    data[i, ] <- currentData
  }
  
  out <- list(
    data = data,
    spacings = spacings,
    x = x,
    N = n,
    samples = samples,
    nEff = nEff,
    alpha = alpha,
    cutoff = cutoff
  )
  
  return(out)
}
```




# Comparisons

Text here.

```{r samplesSetup}
x1 <- c(0.11, 0.19, 0.27, 0.39, 0.42, 0.48, 0.59, 0.71, 0.84, 0.92)
x2 <- c(0.07, 0.12, 0.19, 0.26, 0.31, 0.36, 0.41, 0.49, 0.67, 0.89)
```

```{r popSetup}
N <- 39
nEff <- 4
alpha <- 0.5
cutoff <- 0.5
```

```{r initRuns, include = FALSE}
initRun1 <- importanceSampler(x = x1, N = N, samples = 100, nEff = 4)
initRun2 <- mcmcUniform(x = x1, N = N, samples = 100, nEff = 4)
initRun3 <- mcmcPrior(x = x1, N = N, samples = 100, nEff = 4)

initRun4 <- importanceSampler(x = x2, N = N, samples = 100, nEff = 4)
initRun5 <- mcmcUniform(x = x2, N = N, samples = 100, nEff = 4)
initRun6 <- mcmcPrior(x = x2, N = N, samples = 100, nEff = 4)
```

```{r timings, include = FALSE, cache = TRUE}
time_1 <- profileExpr(pops1_1 <- importanceSampler(x = x1, N = N, samples = samples, nEff = 4))
time_2 <- profileExpr(pops1_2 <- mcmcUniform(x = x1, N = N, samples = samples, nEff = 4))
time_3 <- profileExpr(pops1_3 <- mcmcPrior(x = x1, N = N, samples = samples, nEff = 4))

pops2_1 <- importanceSampler(x = x2, N = N, samples = samples, nEff = 4)
pops2_2 <- mcmcUniform(x = x2, N = N, samples = samples, nEff = 4)
pops2_3 <- mcmcPrior(x = x2, N = N, samples = samples, nEff = 4)

prof_1 <- slice_max(data.frame(flatProfile(time_1, byTotal = TRUE, GC = FALSE)), total.time, n = 10, with_ties = FALSE)
prof_2 <- slice_max(data.frame(flatProfile(time_2, byTotal = TRUE, GC = FALSE)), total.time, n = 10, with_ties = FALSE)
prof_3 <- slice_max(data.frame(flatProfile(time_3, byTotal = TRUE, GC = FALSE)), total.time, n = 10, with_ties = FALSE)
```

## Location (Mean)

```{r means, echo = FALSE, fig.cap = "Density Comparison of Importance Sampling (red), Uniform Proposal MCMC (blue), and Prior Proposal Block MCMC (green)", out.width = "100%"}
plot(density(rowMeans(pops1_1)), col = "red", main = "", xlab = "Mean of Population", ylim = c(0, 7))
lines(density(rowMeans(pops1_2)), col = "blue")
lines(density(rowMeans(pops1_3$data)), col = "green")
```

## Location (CDFs)

```{r cdfs, echo = FALSE, fig.cap = "CDFs of Importance Sampling (left), Uniform Proposal MCMC (center), and Prior Proposal Block MCMC (right)", out.width = "100%", fig.height = 3}
par(mfrow = c(1, 3), mar = c(3, 0, 0, 0))

pops1_1_sort = apply(pops1_1, 2, sort)
plot(c(0, 1), c(0, 1), type = 'l', xlab = 'U', ylab = 'Cumulative Probabilities', xaxt = 'n')
matlines(t(pops1_1_sort), cumsum(rep(1/N, N)), lty = 1, type = 's')
axis(1, 1:9 / 10)
mtext("Importance Sampling")

pops1_2_sort = apply(pops1_2, 2, sort)
plot(c(0, 1), c(0, 1), type = 'l', xlab = 'U', ylab = 'Cumulative Probabilities', xaxt = 'n', yaxt = 'n', ann = FALSE)
matlines(t(pops1_2_sort), cumsum(rep(1/N, N)), lty = 1, type = 's')
axis(1, 1:9 / 10)
mtext("Markov Chains\n(Uniform Proposal)")

pops1_1_sort = apply(pops1_3$data, 2, sort)
plot(c(0, 1), c(0, 1), type = 'l', xlab = 'U', ylab = 'Cumulative Probabilities', xaxt = 'n', yaxt = 'n', ann = FALSE)
matlines(t(pops1_1_sort), cumsum(rep(1/N, N)), lty = 1, type = 's')
axis(1, 1:9 / 10)
mtext("Block Markov Chains\n(Prior Proposal)")
```

## Time

```{r, echo = FALSE}
kables(
  list(
    kable(prof_1, col.names = c("Total %", "Total Time", "Self %", "Self Time")) %>%
      kable_styling(full_width = FALSE, position = "float_left"),
    kable(prof_2, col.names = c("Total %", "Total Time", "Self %", "Self Time")) %>%
      kable_styling(full_width = FALSE, position = "center"),
    kable(prof_3, col.names = c("Total %", "Total Time", "Self %", "Self Time")) %>%
      kable_styling(full_width = FALSE, position = "float_right")
  ),
  caption = "Profile for Importance Sampling (left), Uniform Proposal MCMC (center), and Prior Proposal Block MCMC (right)"
) %>%
  kable_styling(full_width = TRUE, font_size = 11)
```

